Improving the communication and accessibility of stock assessment using interactive

### 2019-04-10 analysis/paper/manuscript.Rmd
Revise the open stock assessment section to clue up the next draft

from the continuation of long-term monitoring efforts or from the addition of new monitoring programs. As such, stock assessment biologists often have to manage large volumes of data from a variety of sources. Time series of reported landings and catch-at-age are often analyzed in conjunction with data from “fishery-independent” surveys that track changes in abundance and, in many cases, also monitor trends in biological factors such as age composition, growth rates, sex ratios and maturation stages. For some data-rich stocks, mark and recapture studies are also carried out to estimate movement, migration, growth rate, natural mortality, and discard mortality. All of the above-mentioned data sets are multidimensional and as the volume and variety of these data increases, it becomes more difficult to be aware of the details of each data set and discover key patterns within each. This challenge can be mediated, to a degree, by the application of interactive visualization tools as that they allow detailed exploration of the data behind the plot. For instance, the ability to zoom in on features or areas of interest and hover over specific points to reveal more information creates an interactive user-driven experience that expedites explorations of the data. This is exemplified by an interactive mapping tool developed for the exploration of a long-term tagging study.

@@

The NCAM dashboard, which is included as a self-contained html file in Supplement 2, contains a series of tabs, the first of which provides terse point-form background on the model (tab named "Background"). Subsequent tabs provide a series of diagnostic plots for assessing model fits to the catch ("Catch"), research vessel autumn trawl survey ("RV survey"), Sentinel fishery survey ("SN survey"), inshore acoustic survey ("SS survey"), and tagging ("Tagging") data. For instance, the "RV survey" tab includes plots of observed and predicted values of mean numbers per tow captured in the research vessel survey (Figure \@ref(fig:rv)). The dashboard also includes tabs focused on model estimates such as catchability and selectivity ("Catchability"), stock size and vital rates ("Trends"; Figure \@ref(fig:trends)), and stock productivity ("Productivity"). Finally, some results from a retrospective analyses are included under the "Retro" tab and key inputs and outputs are included as tables under the "Tables" tab. The plots and tables included in the dashboard are similar to those typically presented at assessment meetings and in research documents, however, the interactive nature of the dashboard permits easy and efficient access to significantly more detail. For instance, the ability to zoom allows users to investigate specific patterns and tooltip display the values behind specific points. Because the dashboard is contained in an html file, it acts as an interactive assessment document that can be shared for peer-review. This allows colleagues and stakeholders to independently scrutinize details of the data and model that are not easily accessed by users other than the analyst. Such access improves the transparency of the stock assessment model which, in turn, leads to richer discussion and scrutiny of the biological and statistical rigor of the model.

> TODO: should I add detail of questions not previously raised? If so, I can't quite remember.

@@

# Open stock-assessment

The amount of data available to scientists has grown by orders of magnitude in recent decades as has the complexities of data management, exploratory data analysis, as well as the formal analyses and associated diagnostics. The majority of this sequence of events, sometimes called “the data pipeline”, have not traditionally been part of the peer-review process which sees only the end products. However, decisions made at these steps increasingly influence the outcome of the study. Gelman coined the term “the forked garden path” to illustrate that different conclusions can be arrived at depending on what decisions one makes along the garden path. Due to a number of limitations, such as available pages in journals, much of the data pipeline is not transparent nor is it reproducible, even if it is one’s own analysis. A number of authors have recently advocated for a culture of open science and reproducible research, i.e., a change in the transparency and reproducibility of science. Proponents of open science and reproducible research posit a number of benefits including a more productive and responsible scientific culture, an ability to address larger and more complex questions, as well as a more efficient workflow and ability to reproduce one’s own work [@lewis2018].

We submit that the dashboard tool is also a major step forward in terms of open science and reproducible research in addition to being a major step forward in how stock assessments are presented and critiqued. Consider the following process. Through the Canadian Science Advisory Secretariat (CSAS), DFO Science provides peer-reviewed science advice to fisheries managers at regular meetings (Regional Assessment Process). These meetings are attended by a variety of stakeholders including harvesters, academics, and NGOs. Documents summarizing these processes (research documents, stock assessment reports, and proceedings summaries) are available to the public through the CSAS website. Ergo the CSAS process strives to uphold the idea of transparency. However, as it currently stands, only the end products of an assessment are open and transparent; as with virtually all scientific analyses, the data pipeline of stock assessment is hidden from review.

The dashboard removes much of this problem by providing access to all model inputs, outputs and diagnostics in an open and easy to use interactive document. In terms of data management, this point is largely ameliorated because the dashboard tool is constructed in Program R using R Markdown and the code for creating the dashboard and all associated analyses are available on Github. Data are also available through information requests.  So the entire data pipeline is transparent as well as reproducible, i.e., the requirements of open science are fulfilled.

The dashboard also promotes accessibility to the stock assessment. Even if one is able to download code from a content management system (e.g., GitHub or Bitbucket), submit a data request, and manipulate the data so that it conforms to the code, it is unlikely that an individual has the time to fully recreate a particular analysis in its entirety when most reviewers are already overcommitted [@banks2011]. In the case of NCAM, considerable statistical experience and expertise is required simply to run the model (or perhaps to even tweak the model). However, the dashboard removes many of these difficulties. The html file contains all of the relevant analyses. The number of analyses can be expanded upon request and can be distributed prior to meetings and reviewers can assess the results and diagnostics at their leisure.

We submit that the dashboard tool is an important step forward in transparent and reproducible science. Further, this tool makes stock assessments more accessible to stakeholders. Beyond stock assessments, interactive documents could be used for a more fulsome review of manuscripts (i.e., reviewers can now assess claims like “we examined the residuals and found that the assumptions of the model were met”) in many fields.

>>>>

The data sets used in stock assessments are constantly growing. This growth in data either stems from the continuation of long-term monitoring efforts or from the addition of new monitoring programs. As such, stock assessment biologists often have to manage large volumes of data from a variety of sources. Time series of reported landings and catch-at-age are often analyzed in conjunction with data from “fishery-independent” surveys that track changes in abundance and, in many cases, also monitor trends in biological factors such as age composition, growth rates, sex ratios and maturation stages. For some data-rich stocks, mark and recapture studies are also carried out to estimate movement, migration, growth rate, natural mortality, and discard mortality. All of the above-mentioned data sets are multidimensional and as the volume and variety of these data increases, it becomes more difficult to be aware of the details of each data set and discover key patterns within each. This challenge can be mediated, to a degree, by the application of interactive visualization tools as that they allow detailed exploration of the data behind the plot. For instance, the ability to zoom in on features or areas of interest and hover over specific points to reveal more information creates an interactive user-driven experience that expedites explorations of the data. This is exemplified by an interactive mapping tool developed for the exploration of a long-term tagging study.

@@

The NCAM dashboard, which is included as a self-contained html file in Supplement 2, contains a series of tabs, the first of which provides terse point-form background on the model (tab named "Background"). Subsequent tabs provide a series of diagnostic plots for assessing model fits to the catch ("Catch"), research vessel autumn trawl survey ("RV survey"), Sentinel fishery survey ("SN survey"), inshore acoustic survey ("SS survey"), and tagging ("Tagging") data. For instance, the "RV survey" tab includes plots of observed and predicted values of mean numbers per tow captured in the research vessel survey (Figure \@ref(fig:rv)). The dashboard also includes tabs focused on model estimates such as catchability and selectivity ("Catchability"), stock size and vital rates ("Trends"; Figure \@ref(fig:trends)), and stock productivity ("Productivity"). Finally, some results from a retrospective analyses are included under the "Retro" tab and key inputs and outputs are included as tables under the "Tables" tab. The plots and tables included in the dashboard are similar to those typically presented at assessment meetings and in research documents, however, the interactive nature of the dashboard permits easy and efficient access to significantly more detail. For instance, the ability to zoom allows users to investigate specific patterns and small boxes, called tooltips, display more information about the points users hover over. Because the dashboard is contained in an html file, it acts as an interactive assessment document that can be shared for peer-review. This allows colleagues and stakeholders to independently scrutinize details of the data and model that are not easily accessed by users other than the analyst. Such access improves the transparency of the stock assessment model which, in turn, leads to richer discussion and scrutiny of the biological and statistical rigor of the model.

@@

# Towards open stock-assessment

The amount of data available to scientists has grown by orders of magnitude in recent decades as has the complexities of data management, exploratory data analysis, as well as the formal analyses and associated diagnostics. The majority of this sequence of events, sometimes called “the data pipeline”, have not traditionally been part of the peer-review process which sees only the end products. However, decisions made at these steps increasingly influence the outcome of the study. Gelman and Loken [-@gelman2014] coined the term "garden of forking paths" to illustrate that different conclusions can be arrived at depending on what decisions one makes along the garden path. Due to a number of limitations, such as available pages in journals, much of the data pipeline is not transparent nor is it reproducible, even if it is one’s own analysis. A number of authors have recently advocated for a culture of open science and reproducible research, i.e., a change in the transparency and reproducibility of science. Proponents of open science and reproducible research posit a number of benefits including a more productive and responsible scientific culture, an ability to address larger and more complex questions, as well as a more efficient workflow and ability to reproduce one’s own work [@lewis2018].

We submit that the dashboard tool is also a step forward in terms of open science and reproducible research in addition to being a major step forward in how stock assessments are presented and critiqued. Consider the following process. Through the Canadian Science Advisory Secretariat (CSAS), DFO Science provides peer-reviewed science advice to fisheries managers at regular meetings (Regional Assessment Process). These meetings are attended by a variety of stakeholders including harvesters, academics, and NGOs. All participants attending a CSAS meeting are encouraged to question, comment and constructively challenge the science presented and the goal is to provide the best possible science advice to the Minister, managers, stakeholders and the public. While the entire data-pipeline can be revealed to all by sharing code and data, it is unlikely that an individual has the time to fully review and recreate a particular analysis in its entirety when most participants are already overcommitted [@banks2011]. In the case of NCAM, considerable statistical experience and expertise is required simply to run the model (or perhaps to even tweak the model). However, the dashboard removes many of these difficulties. The html file contains all model inputs, outputs and diagnostics in an open and easy to use interactive document. The number of analyses can be expanded upon request and can be distributed prior to meetings and participants can assess the results and diagnostics at their leisure. Specific experience with the tools used to generate the results are not required to constructively and critically review of the results presented in the dashboard.

We must acknowledge there are costs to adopting this approach of using the interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively to develop these tools. In our two examples, the developers were relatively proficient with R and have natural aptitudes for programming, but had their formal training in field-based population ecology. Most programming issues were readily addressed through simple internet searchers, but some learning time was certainly needed (a few weeks). As more people in the fisheries community develop and use these tools, we expect tool develop speed and quality will only increase. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding. Scientific exploration is rarely a linear process, and the ability to jump around a dashboard is a more in tune with the natural way scientist’s pursue ideas. In contrast, a slide presentation is effectively a linear and a single way of developing a scientific narrative; a method hardly conducive to exploring complex data problems.

In summary, we submit that the dashboard tool is an important step forward in transparent and reproducible science. Further, this tool makes stock assessments more accessible to stakeholders. Beyond stock assessments, interactive documents could be used for a more fulsome review of manuscripts (i.e., reviewers can now assess claims like “we examined the residuals and found that the assumptions of the model were met”) in many fields.
